---
title: "Data Visualization"
subtitle: "Visualizing 2D categorical and continuous by categorical"
date: "June 4th, 2025"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    # chakra: "libs/remark-latest.min.js"
    # css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    self-contained: yes
    # css: [default, default-fonts]
    # seal: false
    # anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options:
  chunk_output_type: console
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#cc002b")
```

```{r setup, echo = FALSE}
knitr::opts_chunk$set(fig.retina = 3, warning = FALSE, message = FALSE)
```

```{r get-ohtani, warning = FALSE, message = FALSE}
library(baseballr)
```
## Revisiting MVP Shohei Ohtani's batted balls in 2021

Created dataset of batted balls by the American League MVP Shohei Ohtani in 2021 season using [`baseballr`](http://billpetti.github.io/baseballr/)

```{r load-ohtani, warning = FALSE, message = FALSE}
library(tidyverse)
ohtani_batted_balls <- read_csv("https://shorturl.at/mnwL1")
head(ohtani_batted_balls)
```

- each row / observation is a batted ball from Ohtani's 2021 season
- __Categorical__ / qualitative variables: `pitch_type`, `batted_ball_type`, `outcome`
- __Continuous__ / quantitative variables: `hit_x`, `hit_y`, `exit_velocity`, `launch_angle`

---

## First - more fun with [`forcats`](https://forcats.tidyverse.org/)

Variables of interest: [`pitch_type`](https://library.fangraphs.com/pitch-type-abbreviations-classifications/) and `batted_ball_type` - but how many levels does `pitch_type` have?

```{r pitch-type-table}
table(ohtani_batted_balls$pitch_type)
```

We can manually [`fct_recode`](https://forcats.tidyverse.org/reference/fct_recode.html) `pitch_type` (see [Chapter 15 of `R` for Data Science](https://r4ds.had.co.nz/factors.html) for more on factors)

```{r recode-pitch-type}
ohtani_batted_balls <- ohtani_batted_balls %>%
  filter(pitch_type != "null") %>% 
  mutate(pitch_type = fct_recode(pitch_type, "Changeup" = "CH", "Breaking ball" = "CU", #<<
                      "Fastball" = "FC", "Fastball" = "FF", "Fastball" = "FS", #<<
                      "Breaking ball" = "KC",  "Fastball" = "SI",  "Breaking ball" = "SL")) #<<
```
__Question:__ Are all pitch types equally likely to occur?
---

## Inference for categorical data

The main test used for categorical data is the __chi-square test__:

  - __Null hypothesis__: $H_0: p_1 = p_2 = \cdots = p_K$ and we compute the __test statistic__: 

$$
\chi^2 = \sum_{j=1}^K \frac{(O_j - E_j)^2}{E_j}
$$

  - $O_j$: observed counts in category $j$

  - $E_j$: expected counts under $H_0$ (i.e., $\frac{n}{K}$ or each category is equally likely to occur)

--

```{r}
chisq.test(table(ohtani_batted_balls$pitch_type)) #<<
```


---

## Statistical inference in general

Computing $p$-values works like this:

- Choose a test statistic.

- Compute the test statistic in your dataset.

- Is test statistic "unusual" compared to what I would expect under $H_0$?

- Compare $p$-value to __target error rate__ $\alpha$ (typically referred to as target level $\alpha$ )

- Typically choose $\alpha = 0.05$ 


---

## 2D Categorical visualization (== more bar charts!)

.pull-left[

__Stacked__: a bar chart of _spine_ charts

```{r stacked-bars, fig.height=4}
ohtani_batted_balls %>%
  ggplot(aes(x = batted_ball_type,
             fill = pitch_type)) + #<<
  geom_bar() + theme_bw()
```


]
.pull-right[

__Side-by-Side__: a bar chart _of bar charts_

```{r side-by-side-bars, fig.height=4}
ohtani_batted_balls %>%
  ggplot(aes(x = batted_ball_type,
             fill = pitch_type)) + 
  geom_bar(position = "dodge") + theme_bw() #<<
```
]

---

## Which do you prefer?

.pull-left[

```{r ref.label='stacked-bars', fig.height=4, echo = FALSE}

```


]
.pull-right[

```{r, ref.label='side-by-side-bars', fig.height=4, echo = FALSE}

```
]

--


- Stacked bar charts emphasize __marginal__ distribution of `x` variable, 
  - e.g. $P$ (`batted_ball_type` = fly_ball)

- Side-by-side bar charts are useful to show the __conditional__ distribution of `fill` variable given `x`,
  - e.g. $P$ (`pitch_type` = Fastball | `batted_ball_type` = fly_ball)


---

## Contingency tables

Can provide `table()` with more than one variable

```{r}
table("Pitch type" = ohtani_batted_balls$pitch_type, 
      "Batted ball type" = ohtani_batted_balls$batted_ball_type)
```

--

Easily compute `proportions()`:

```{r}
proportions(table(ohtani_batted_balls$pitch_type, ohtani_batted_balls$batted_ball_type)) #<<
```

---

## Review of joint, marginal, and conditional probabilities

__Joint distribution__: frequency of intersection, $P(X = x, Y = y)$

```{r}
proportions(table(ohtani_batted_balls$pitch_type, ohtani_batted_balls$batted_ball_type))
```

--

__Marginal distribution__: row / column sums, e.g. $P(X = \text{popup}) = \sum_{y \in \text{pitch types}} P(X = \text{popup}, Y = y)$

--

__Conditional distribution__: probability event $X$ __given__ second event $Y$, 
- e.g. $P(X = \text{popup} | Y = \text{Fastball}) = \frac{P(X = \text{popup}, Y = \text{Fastball})}{P(Y = \text{Fastball})}$

---

### BONUS: `pivot_wider` example

Manually construct this table for practice...

```{r joint-table}
library(gt)
ohtani_batted_balls %>%
  group_by(batted_ball_type, pitch_type) %>%
  summarize(joint_prob = n() / nrow(ohtani_batted_balls)) %>%
  pivot_wider(names_from = batted_ball_type, values_from = joint_prob,
              values_fill = 0) %>%
  gt()
```


---

## Inference for 2D categorical data

We AGAIN use the __chi-square test__:

- __Null hypothesis__: $H_0$: Variables $A$ and $B$ are independent, 

  - e.g., `batted_ball_type` and `pitch_type` are independent of each other, no relationship

--

- And now we compute the __test statistic__ as: 

$$\chi^2 = \sum_i^{k_1} \sum_j^{k_2} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

- $O_{ij}$: observed counts in contingency table $j$

- $E_{ij}$: expected counts under $H_0$ where __under the null__:

$$
\begin{aligned}
E_{ij} &= n \cdot P(A = a_i, B = b_j) \\
&= n \cdot P(A = a_i) P(B = b_j) \\
&= n \cdot \left( \frac{n_{i \cdot}}{n} \right) \left( \frac{ n_{\cdot j}}{n} \right)
\end{aligned}
$$
  
---

## Inference for 2D categorical data

We AGAIN use the __chi-square test__:

- __Null hypothesis__: $H_0$: Variables $A$ and $B$ are independent, 

  - e.g., `batted_ball_type` and `pitch_type` are independent of each other, no relationship

- And now we compute the __test statistic__ as: 

$$\chi^2 = \sum_i^{k_1} \sum_j^{k_2} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

```{r}
chisq.test(table(ohtani_batted_balls$pitch_type, ohtani_batted_balls$batted_ball_type)) #<<
```


---

## Can we visualize independence?

--
Two variables are __independent__ if knowing the level of one tells us nothing about the other
- i.e.  $P(X = x | Y = y) = P(X = x)$, and that $P(X = x, Y = y) = P(X = x) \times P(Y = y)$

--

.pull-left[

Create a __mosaic__ plot using __base `R`__

```{r mosaic, eval = FALSE}
mosaicplot(table(ohtani_batted_balls$pitch_type, ohtani_batted_balls$batted_ball_type),
           main = "Relationship between batted ball and pitch type?")
```

- spine chart _of spine charts_

- width $\propto$ marginal distribution of `pitch_type`

- height $\propto$ conditional distribution of `batted_ball_type` | `pitch_type`

- area $\propto$ joint distribution

__[`ggmosaic`](https://github.com/haleyjeppson/ggmosaic) has issues...__
]
.pull-right[
```{r ref.label = 'mosaic', echo = FALSE, fig.height=6}

```
]

---

## Shade by _Pearson residuals_

- The __test statistic__ is: 

$$\chi^2 = \sum_i^{k_1} \sum_j^{k_2} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

- Define the _Pearson residuals_ as:

$$r_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij}}}$$

- Sidenote: In general, Pearson residuals are $\frac{\text{residuals}}{\sqrt{\text{variance}}}$

--

- $r_{ij} \approx 0 \rightarrow$ observed counts are close to expected counts

- $|r_{ij}| > 2 \rightarrow$ "significant" at level $\alpha = 0.05$.

--

- Very positive $r_{ij} \rightarrow$ more than expected, while very negative $r_{ij} \rightarrow$ fewer than expected

- Mosaic plots: Color by Pearson residuals to tell us which combos are much bigger/smaller than expected.

---

## Shade by _Pearson residuals_

```{r mosaic-shade, fig.align='center'}
mosaicplot(table(ohtani_batted_balls$pitch_type, ohtani_batted_balls$batted_ball_type),
           shade = TRUE, main = "Relationship between batted ball and pitch type?") #<<
```

---

## Continuous by categorical: side-by-side and color

.pull-left[

```{r violin-box, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(x = pitch_type, #<<
             y = exit_velocity)) +
  geom_violin() +
  geom_boxplot(width = .2) +
  theme_bw()
```

```{r ref.label = 'violin-box', echo = FALSE, fig.height=4}
```
  
]
.pull-right[
```{r ecdf, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity,
             color = pitch_type)) + #<<
  stat_ecdf() + 
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r ref.label = 'ecdf', echo = FALSE, fig.height=4}
```
]

---

## What about for histograms? 

.pull-left[

```{r stacked-hist, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity,
             fill = pitch_type)) + #<<
  geom_histogram() +
  theme_bw() + theme(legend.position = "bottom")
```

```{r ref.label = 'stacked-hist', echo = FALSE, fig.height=4}
```
  
]
.pull-right[
```{r overlaid-hist, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity,
             fill = pitch_type)) + 
  geom_histogram(alpha = .25, position = "identity") + #<<
  theme_bw() + theme(legend.position = "bottom")
```

```{r ref.label = 'overlaid-hist', echo = FALSE, fig.height=4}
```
]

---

## We can always facet instead...

.pull-left[

```{r facet-hist-wrap, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity)) + 
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ pitch_type, ncol = 2) #<<
```

```{r ref.label = 'facet-hist-wrap', echo = FALSE, fig.height=4}
```
  
]
.pull-right[
```{r facet-hist-grid, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity)) + 
  geom_histogram() +
  theme_bw() +
  facet_grid(pitch_type ~., margins = TRUE) #<<
```

```{r ref.label = 'facet-hist-grid', echo = FALSE, fig.height=4}
```
]

---

## Facets make it easy to move beyond 2D

```{r stacked-bars-facet, fig.width=12, fig.height=5, fig.align='center'}
ohtani_batted_balls %>%
  ggplot(aes(x = pitch_type,
             fill = batted_ball_type)) + 
  geom_bar() + theme_bw() +
  facet_wrap(~ outcome, ncol = 5) +
  theme(legend.position = "bottom")
```


---

## 2D Continuous Relationships --> Scatterplot

.pull-left[

- We make a __scatterplot__ with [`geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html)


```{r scatter, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity, 
             y = launch_angle)) + 
  geom_point() + 
  theme_bw()
```
  
]
.pull-right[
```{r ref.label = 'scatter', echo = FALSE, fig.height=7}
```
]

---

## Two continuous, one categorical...

.pull-left[



```{r scatter2, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity, 
             y = launch_angle,
             color = batted_ball_type)) +  #<<
  geom_point() + 
  theme_bw()
```
The possibilities are endless!

]

.pull-right[
```{r ref.label = 'scatter2', echo = FALSE, fig.height=7}
```
]



