---
title: "Lab 9 - Bootstrapping"
output: html_notebook
---

```{r}

library(tidyverse)
library(ggplot2)

red = 21
n = 21 + 26 + 21 + 24 + 22 

red_vec <- c(rep(TRUE, red), rep(FALSE, n - red))

# Shuffle the vector
red_vec <- sample(red_vec)

# Create data frame
mnms <- data.frame(entry = 1:n, red = red_vec)

head(mnms)

```
```{r}
#construct confidence intervals
red_percent = red/n

wald_se = sqrt((red_percent * (1 - red_percent)) / n)
wald_lower = red_percent - 1.96 * wald_se
wald_upper = red_percent + 1.96 * wald_se

wald_df <- data.frame(
  proportion = red_percent,
  lower = wald_lower,
  upper = wald_upper
  )

n_tilde = n + 4
p_tilde = (red + 2) / n_tilde
ac_se = sqrt((p_tilde * (1 - p_tilde)) / n_tilde)
ac_lower = p_tilde - 1.96 * ac_se
ac_upper = p_tilde + 1.96 * ac_se

ac_df <- data.frame(
  proportion = red_percent,
  lower = ac_lower,
  upper = ac_upper
  )

ggplot(wald_df, aes(x = 1, y = proportion)) +
  geom_point(color = "darksalmon", size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_hline(yintercept = red_percent, linetype = "dashed", color = "red") +
  ylim(0, 1) +
  labs(
    title = "Wald 95% Confidence Interval for Proportion Red",
    y = "Proportion Red",
    x = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())


```
```{r}
#bootstrap 

set.seed(9)
permutations = 100000 

bootstrap_p <- numeric(permutations)

for (i in 1:permutations) {
  bootstrap_sample <- sample(mnms$red, replace = TRUE)
  p <- sum(bootstrap_sample) / length(bootstrap_sample)
  bootstrap_p[i] <- p
}

# Save into a data frame
bootstrap_results <- data.frame(p = bootstrap_p)

# Preview
head(bootstrap_results)

ggplot(bootstrap_results, aes(x = p)) +
  geom_histogram(binwidth = 0.01, fill = "#FF6B6B", color = "white", boundary = 0) +
  labs(
    title = "Bootstrap Distribution of Proportion Red",
    x = "Proportion Red (p)",
    y = "Frequency"
  ) +
  theme_minimal()
ci <- quantile(bootstrap_results$p, probs = c(0.025, 0.975))

ci_df <- tibble::tibble(
  method = c("Wald", "Agresti-Coull", "Bootstrap"),
  lower = c(wald_lower, ac_lower, ci[1]),
  upper = c(wald_upper, ac_upper, ci[2]),
  estimate = red_percent
)

ggplot(ci_df, aes(x = method, y = estimate)) +
  geom_point(color = "red", size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.05, color = "black") +
  ylim(0, 0.4) +
  labs(
    title = "95% Confidence Intervals for Proportion Red",
    y = "Proportion",
    x = "Method"
  ) +
  theme_minimal()

```

We see that the confidence intervals are relatively similar across the methods, with wald and bootstrap being almost identical. The agresti-coull interval looks to be slightly larger but it's difficult to tell due to the additional positive transformation of the interval.


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

