---
title: "12_Maximilian"
author: "Maximilian J. Gebauer"
date: "2025-06-18"
output: html_document
---

```{r Setup, include=FALSE, results='hide', warning=FALSE}
knitr::opts_chunk$set(echo = T, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output

# Package setup
if(!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, dplyr, ggthemes, data.table, lubridate, glmnet,
               GGally, RColorBrewer, ggsci, plotROC, usmap,
               plotly, ggpubr, vistime, coefplot, skimr, car, ggrepel, slider, lubridate,
               tidymodels,ranger,vip,ggplot2, tune,dials,pdp, purrr, stringr, lmtest,
               sandwich)
```


posterior mean of mu_hat_i,1 (one game data) is given by (((Xij/Pij)/sigma2) + (1/tau2) x mu_global)) / (the sum of the between game precision for player and between player precision)

for j>2, posterior mean of mu_hat_j,i is given by: (((Xij/Pij x N_j)/sigma2) + (1/tau2) x mu_global)) / (the sum of the between player precision and the prior precision given by 1 over the variane of mu_i,j-1)

```{r}
nba_data <- read_csv("/Users/maximiliangebauer/Documents/summer-lab/2025/labs/data/12_nba-box-scores.csv")
head(nba_data)
```
```{r}
sub_nba <- nba_data %>%
  group_by(idPlayer) %>%
  mutate(N=n()) %>%
  filter(N > 9) %>%
  ungroup()

sub_nba <- sub_nba %>%
  filter(possessions > 0)
```

```{r}
player_results <- sub_nba %>%
  group_by(idPlayer) %>%
  mutate(
    N = n(),
    mean_ppp = mean(pts / possessions),
    var_ppp = var(pts / possessions),
  ) %>%
  select(idPlayer,namePlayer,mean_ppp,var_ppp,N) %>%
  unique() %>%
  ungroup

head(player_results)
```

```{r}
(mu_hat <- mean(player_results$mean_ppp))
(var_hat <- var(player_results$mean_ppp))
```

```{r}
tolerance_upper <- mu_hat + sqrt(var_hat)
tolerance_lower <- mu_hat - sqrt(var_hat)

filt_players <- player_results %>%
  filter(mean_ppp <= tolerance_upper & mean_ppp >= tolerance_lower)


(mean_vars <- mean(filt_players$var_ppp))
```

```{r}
df <- sub_nba

split_df <- df %>%
  group_by(idPlayer) %>%
  mutate(
    row   = row_number(),
    total = n(),
    sum_pos = sum(possessions),
    sum_pts = sum(pts),
    is_train = row <= ceiling(total/2)
  ) %>%
  ungroup()

train_df <- split_df %>%
  filter(is_train) %>%
  select(-row, -total, -is_train)

test_df <- split_df %>%
  filter(!is_train) %>%
  select(-row, -total, -is_train)

head(train_df)
```
```{r}
train_results <- train_df %>%
  group_by(idPlayer) %>%
  mutate(
    mean_ppp = mean(pts / possessions),
    var_ppp = var(pts / possessions),
    N = n()
  ) %>%
  select(idPlayer,namePlayer,mean_ppp,var_ppp,N) %>%
  unique() %>%
  ungroup

(mu_hat_train <- mean(train_results$mean_ppp))
(var_hat_train <- var(train_results$mean_ppp))
```

```{r}
test_results <- test_df %>%
  group_by(idPlayer) %>%
  mutate(
    mean_g2_ppp = mean(pts / possessions),
  ) %>%
  select(idPlayer,mean_g2_ppp) %>%
  unique() %>%
  ungroup

total_results <- train_results %>%
  left_join(test_results,by="idPlayer")

head(total_results)
```



```{r}
global_mu    <- mu_hat_train
tau2_seq     <- seq(1e-6, 1e-2, length.out = 100)
n_players    <- nrow(total_results)
rmse_vec     <- numeric(length(tau2_seq))

#sigma2_train <- sigma_hat_train^2
#total_results$var_ppp[is.na(total_results$var_ppp)] <-
  #sigma2_train / total_results$N[is.na(total_results$var_ppp)]

for (i in seq_along(tau2_seq)) {
  tau2   <- tau2_seq[i]
  mu_pred <- numeric(n_players)
  
  for (j in seq_len(n_players)) {
    nj     <- total_results$N[j]
    ybarj  <- total_results$mean_ppp[j]
    vj     <- total_results$var_ppp[j]
    
    # posterior precision for ȳ_j
    prec_y <- nj / vj
    prec_0 <- 1  / tau2
    
    mu_pred[j] <- ( prec_y * ybarj + prec_0 * global_mu ) /
                  ( prec_y + prec_0 )
  }
  
  # RMSE across all players for this tau2
  rmse_vec[i] <- sqrt(
    mean((mu_pred - total_results$mean_g2_ppp)^2, na.rm = TRUE)
  )
}

results <- data.frame(tau2 = tau2_seq, rmse = rmse_vec)
```

```{r}
min_rmse <- min(results$rmse)

argmin_tau2 <- results %>%
  filter(rmse == min_rmse) %>%
  select(tau2)

argmin_tau2 <- argmin_tau2$tau2

plot(results$tau2, results$rmse, type = "l",
     xlab = expression(tau^2), ylab = "RMSE",
     main = "RMSE vs. tau^2")
abline(v = argmin_tau2,      # the x-intercept
       col = "red",    # line color
       lty = 2,        # dashed line type
       lwd = 2)

argmin_tau2
min_rmse
```

optimum tau2 for minimizing rmse is 0.000809

```{r}
task6 <- sub_nba %>%
  arrange(idPlayer, idGame)

task6 <- task6 %>%
  mutate(
    ppp = pts / possessions,
  )
head(task6)
```

```{r}
sigma2 <- 0.01
# 2. Compute running (cumulative) posterior mean after each game
post_traj <- task6 %>%
  group_by(idPlayer) %>%
  mutate(
    n_obs    = row_number(),           # 1,2,3,… games seen so far
    cum_mean = cummean(ppp),           # running sample mean of ppp
    post_mu  = ( (n_obs/sigma2)*cum_mean + (1/tau2)*global_mu ) /
               ( (n_obs/sigma2)        + (1/tau2) )
  ) %>%
  ungroup()

# 3. Pick a random sample of players to plot 
set.seed(42)
sample_players <- sample(unique(post_traj$idPlayer), 6)

post_traj %>%
  filter(idPlayer %in% sample_players) %>%
  group_by(idPlayer) %>%
  mutate(game_index = row_number()) %>%    # 1,2,3,… per player
  ungroup() %>%
  ggplot(aes(x = game_index, y = post_mu, color = factor(idPlayer))) +
    geom_line(size = 1) +
    labs(
      x     = "Game Number in Career",
      y     = expression(mu[bij]),
      color = "Player"
    ) +
    scale_x_continuous(breaks = scales::pretty_breaks(10)) +  # nicer tick placement
    theme_minimal()
```



```{r}
kick_data <- read_csv("/Users/maximiliangebauer/Documents/summer-lab/2025/labs/data/12_field-goals.csv")
head(kick_data)
```
Fit simple logistic model of logit(fg made) on yardline
```{r}
model_1 <- glm(as.factor(fg_made) ~ ydl, data = kick_data, family = "binomial")
summary(model_1)
```

formula from logistic regression: p(make) = exp(4.10758 - 0.10470 x ydl) / (1 + exp(4.10758 - 0.10470 x ydl)), did not use splines or polynomials for simplicitly later on, I would use them if I wanted a better model.



We use the additive recursive nature of the formula for KQ to define a closed form formula for KQ for the ith player prior to/at the jth attempt only in terms of alpha and FGPA. We rely on the fact that KQ_i,0 = 0 (def). Observe that for KQ_i,j = alpha x KQ_i,j-1 + FGPA_i,j-1, we can write KQ_i,j-1 recursively as KQ_i,j-1 = alpha x KQ_i,j-2 + FGPA_i,j-2, iterating this recursive step and observing that base KQ_i,0 = 0, and since the "nesting" operation in the full recursive form is addition, we derive: KQ_i,j = sum from k=1 to j-1 of alpha^(j-k-1) x FGPA_i,k where k indexes kicks, j is the index of the jth kick where we want KQ before that kick occurs, and i indexes Kicker identity. The formula for j=1 (no kicks) returns 0, as it must, for j=2 we have KQ_i,2 = alpha^(2-1-1) x FGPA_i,1 = FGPA_i,1, as it must, and for j=3 we have: alpha x  FGPA_i,1 + FGPA_i,2, as it must. With a for loop we can still use the fully recursive form, but the closed form solution is valid.


```{r}
kicker_data <- kick_data %>%
  group_by(kicker) %>%
  arrange(season, week, .by_group = TRUE) %>%
  ungroup()

kq_update <- function(prev_kq,outcome,alpha, ydl) {
  new_kq = alpha * prev_kq + outcome - exp(4.10758 - 0.10470*ydl) / (1 + exp(4.10758 - 0.10470*ydl))
}
```

```{r}
kickers   <- unique(kicker_data$kicker)
alpha     <- 0.999

# prepare an output column
kicker_data$KQ <- NA_real_

for (pl in kickers) {
  data_i    <- filter(kicker_data, kicker == pl)
  n         <- nrow(data_i)
  
  # if you want to keep the initial zero as element 1, do n+1:
  kq_vec    <- numeric(n + 1)
  kq_vec[1] <- 0
  
  for (j in seq_len(n)) {
    # ### make sure this matches your real column name ###
    made   <- data_i$fg_made[j]   # or data_i$FG[j], etc.
    ydl    <- data_i$ydl[j]
    
    # baseline prob
    p0     <- plogis(4.10758 - 0.10470 * ydl)
    fgpa   <- made - p0
    
    # write into j+1
    kq_vec[j+1] <- alpha * kq_vec[j] + fgpa
  }
  
  # if you only care about the post-kick KQs, drop the first element:
  kicker_data$KQ[kicker_data$kicker == pl] <- kq_vec[-1]
}
```


Plotted kickers N > 300 kicks below (should be ones best represented by the model given selection bias), we see KQ trends are noiser for earlier observations when the sample is small and the kicker is early career, but tend to stabilze more over time. Also tend to increase over time, indicating that these kickers improved over their careers (naturally, we do not observe kickers who get worse consistetently over time with many kicks, as such kickers presumably are cut/do not start). 


```{r}
kicker_data %>%
  group_by(kicker) %>%
  mutate(N=n(),
         kick_index = row_number()) %>%
  filter(N > 300) %>%
  ungroup() %>%
  ggplot(aes(x=kick_index,y=KQ,group=kicker,color=kicker)) +
  geom_point(size=1,alpha=0.5)
```






